---
title: |
  Explainable Artificial Intelligence and Pharmacogenomics for
  Opioid and Polypharmacy Risk Prediction:
  A Dissertation by Publication
subtitle: "Ph.D. in Health Related Sciences (Translational Health Research)"
author: "R. Jerome Dixon, B.S., M.S."
date: "June 2026"

format:
  docx:
    toc: true
    toc-depth: 3
    number-sections: true
  pdf:
    documentclass: book
    pdf-engine: xelatex
    toc: true
    toc-depth: 3
    lot: true
    lof: true
    number-sections: true
    geometry:
      - top=1in
      - bottom=1in
      - left=1.5in
      - right=1in
    fontsize: 12pt
    linestretch: 2
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{multirow}
        \usepackage{makecell}
        \usepackage{xcolor}
        \usepackage{colortbl}
        \usepackage{caption}
        \usepackage{hyperref}
        \hypersetup{hidelinks}
        \usepackage{setspace}
        \usepackage{fancyhdr}
        \usepackage{emptypage}
        \pagestyle{fancy}
        \fancyhead[LE,RO]{\thepage}
        \fancyhead[LO,RE]{\leftmark}
        \fancyfoot{}

bibliography:
  - ../refs/discipline.bib
  - ../refs/bmic-jpm.bib
  - ../refs/cpt-psp.bib
  - ../refs/cts.bib
csl: ../templates/csl/ama.csl

knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

```{r setup}
#| include: false
include_chapter <- function(path) {
  lines <- readLines(path, encoding = "UTF-8")

  # 1. Strip YAML front matter (between first and second "---")
  dashes <- which(trimws(lines) == "---")
  if (length(dashes) >= 2) lines <- lines[(dashes[2] + 1):length(lines)]

  # 2. Replace ```{tikz} chunks with pre-rendered image references
  out <- character()
  in_tikz <- FALSE; lbl <- "fig-unknown"; cap <- ""; wid <- "80%"
  for (ln in lines) {
    if (!in_tikz && grepl("^```\\{tikz", ln)) {
      in_tikz <- TRUE; lbl <- "fig-unknown"; cap <- ""; wid <- "80%"
    } else if (in_tikz && grepl("^```\\s*$", ln)) {
      in_tikz <- FALSE
      fig_base <- tools::file_path_sans_ext(path)
      fig_path <- paste0(fig_base, "_files/figure-pdf/", lbl, "-1.pdf")
      out <- c(out, "",
               paste0("![", cap, "](", fig_path, "){#", lbl,
                      " width=", wid, " fig-align=\"center\"}"),
               "")
    } else if (in_tikz) {
      if (grepl("^%\\| label:", ln))     lbl <- trimws(sub(".*label:\\s*", "", ln))
      if (grepl("^%\\| fig-cap:", ln))   cap <- trimws(sub(".*fig-cap:\\s*", "", ln))
      if (grepl("^%\\| out-width:", ln)) wid <- trimws(sub(".*out-width:\\s*", "", ln))
    } else {
      out <- c(out, ln)
    }
  }

  # 3. Strip trailing References section
  ref_h <- grep("^#{1,6} References\\s*$", out)
  if (length(ref_h)) out <- out[seq_len(tail(ref_h, 1) - 1)]

  cat(out, sep = "\n")
}
```

`\frontmatter`

# Title Page {.unnumbered}

\begin{center}
\vspace*{1.5cm}
{\LARGE\textbf{Explainable Artificial Intelligence and Pharmacogenomics}}\\[0.4em]
{\LARGE\textbf{for Opioid and Polypharmacy Risk Prediction}}\\[1.2em]
{\large A Dissertation by Publication}\\[2em]
Submitted by\\[0.4em]
{\large\textbf{R. Jerome Dixon, B.S., M.S.}}\\[0.3em]
{\normalsize dixonrj@vcu.edu \quad | \quad ORCID: 0000-0001-8622-0597}\\[2em]
{\normalsize Ph.D. Program in Health Related Sciences (Translational Health Research)}\\[0.2em]
{\normalsize Department of Pharmacotherapy and Outcomes Science}\\[0.2em]
{\normalsize School of Pharmacy}\\[0.2em]
{\normalsize Virginia Commonwealth University}\\[0.2em]
{\normalsize Richmond, Virginia}\\[2em]
{\normalsize June 2026}\\[2em]
\textit{Submitted in partial fulfillment of the requirements for the degree of}\\
\textit{Doctor of Philosophy in Health Related Sciences}
\end{center}

\newpage

# Committee Declaration {.unnumbered}

This dissertation, entitled *Explainable Artificial Intelligence and Pharmacogenomics for Opioid and Polypharmacy Risk Prediction*, submitted by **R. Jerome Dixon** in partial fulfillment of the requirements for the degree of Doctor of Philosophy in Health Related Sciences at Virginia Commonwealth University, has been approved by the Committee having charge of the candidate's work.

| Role | Name | Signature | Date |
| :--- | :--- | :--- | :--- |
| Chair | Elvin T. Price, Pharm.D., Ph.D., FAHA | | |
| Member | Tamas Gal, Ph.D. | | |
| Member | Lukasz Kurgan, Ph.D. | | |
| Member | Dayanjan Wijesinghe, Ph.D. | | |
| Member | Jonathan DeShazo, Ph.D. | | |

\newpage

# Publications Included in This Thesis {.unnumbered}

**IRB Protocol:** HM20022300 (non-human-subjects research / IRB waiver; applies to Chapters 3 and 4).

---

**Chapter 1:** R. Jerome Dixon and Elvin T. Price. (2026). *Bridging Explainable Artificial Intelligence and Pharmacogenomics for Opioid and Polypharmacy Risk Prediction: A Systematic Quantitative Literature Review*. Manuscript prepared for submission to *Journal of Personalized Medicine*.

| Contributor | Statement of Contribution | Signature |
| :--- | :--- | :--- |
| R. Jerome Dixon (Candidate) | Conception and design of the research project (95%). Analysis and interpretation of research data (95%). Drafting significant parts of the scholarly work (100%). | |
| Elvin T. Price (Advisor) | Critically revising the work and supervision (10%). | |

---

**Chapter 2:** R. Jerome Dixon and Elvin T. Price. (2026). *Building the Clinical OODA Loop: A Partition-First Data Architecture for Model-Based Precision Therapeutics*. Manuscript prepared for submission to *CPT: Pharmacometrics & Systems Pharmacology*.

| Contributor | Statement of Contribution | Signature |
| :--- | :--- | :--- |
| R. Jerome Dixon (Candidate) | Conception and design of the research project (95%). Data curation and pipeline development (100%). Analysis and interpretation of research data (95%). Drafting significant parts of the scholarly work (100%). | |
| Elvin T. Price (Advisor) | Critically revising the work and supervision (10%). | |

---

**Chapter 3:** R. Jerome Dixon and Elvin T. Price. (2026). *Causal Temporal Drivers of Opioid-Related Emergency Department Visits in Young and Mid-Life Adults: An Ensemble Machine Learning Study with Consensus-Based Feature Attribution*. Manuscript prepared for submission to *Clinical and Translational Science*.

| Contributor | Statement of Contribution | Signature |
| :--- | :--- | :--- |
| R. Jerome Dixon (Candidate) | Conception and design of the research project (95%). Data curation and model development (100%). Analysis and interpretation of research data (95%). Drafting significant parts of the scholarly work (100%). | |
| Elvin T. Price (Advisor) | Critically revising the work and supervision (10%). | |

---

**Chapter 4:** R. Jerome Dixon and Elvin T. Price. (2026). *Formal Feature Attribution as a Causal Calculator for Drug-Drug Interaction Risk in Polypharmacy: A Systems Pharmacology Approach Using All-Payer Claims Data*. Manuscript prepared for submission to *CPT: Pharmacometrics & Systems Pharmacology*.

| Contributor | Statement of Contribution | Signature |
| :--- | :--- | :--- |
| R. Jerome Dixon (Candidate) | Conception and design of the research project (95%). Data curation and model development (100%). Analysis and interpretation of research data (95%). Drafting significant parts of the scholarly work (100%). | |
| Elvin T. Price (Advisor) | Critically revising the work and supervision (10%). | |

---

**Chapter 5:** R. Jerome Dixon and Elvin T. Price. (2026). *A Privacy-First Serverless Pharmacogenomic Risk Dashboard: Translating Ensemble Models and Causal Rules to Clinical Decision Support*. Manuscript prepared for submission to *Journal of Personalized Medicine*.

| Contributor | Statement of Contribution | Signature |
| :--- | :--- | :--- |
| R. Jerome Dixon (Candidate) | Conception and design of the research project (95%). System architecture and implementation (100%). Analysis and interpretation of research data (90%). Drafting significant parts of the scholarly work (100%). | |
| Elvin T. Price (Advisor) | Critically revising the work and supervision (10%). | |

\newpage

# Abstract {.unnumbered}

This dissertation presents a five-study thesis by publication addressing the integration of Explainable Artificial Intelligence (XAI), pharmacogenomics (PGx), and causally valid temporal modeling into a unified, production-deployed clinical decision support framework for opioid and polypharmacy risk prediction from All-Payer Claims Data (APCD).

Chapter 1 establishes the knowledge gap through a Systematic Quantitative Literature Review of 5,839 candidate studies, identifying a distinct lack of research integrating XAI with PGx guidelines at the intersection of opioid and polypharmacy risk modeling. Chapter 2 introduces the foundational partition-first APCD data architecture and the Consensus-Causal Filter combining SHAP and Formal Feature Attribution (FFA), achieving a 15× throughput improvement and strict temporal leakage prevention across 15 Age Band × Cohort partitions. Chapter 3 applies the validated pipeline to predict opioid-related ED visits, demonstrating AUROC ≥ 0.957 and PR-AUC ≥ 0.840 across seven age bands with verifiable DTW-derived trajectory archetypes. Chapter 4 adapts the pipeline as a causal calculator for polypharmacy risk, identifying 115 synergistic drug pairs and 5,021 high-risk triplets via FFA with pharmacogenomically informed Interaction Effect estimation. Chapter 5 translates these models and causal rules into a privacy-first, serverless AWS PGx Risk Dashboard achieving 6 ms warm inference latency (cold-start 2,100 ms) with zero persistent patient data storage. Chapter 6 synthesizes all contributions, evaluates their collective answer to the overarching research aims, and defines a translational roadmap for multi-state federated deployment.

`\mainmatter`

# Chapter 1: Bridging Explainable Artificial Intelligence and Pharmacogenomics for Opioid and Polypharmacy Risk Prediction: A Systematic Quantitative Literature Review {#chap-1}

```{r ch01}
#| results: asis
include_chapter("../CH_1/ch01_cts.qmd")
```

\cleardoublepage

# Chapter 2: Building the Clinical OODA Loop: A Partition-First Data Architecture for Model-Based Precision Therapeutics {#chap-2}

**Foreword to Chapter 2**

The previous chapter established the critical knowledge gap in current clinical predictive modeling through a Systematic Quantitative Literature Review (SQLR), identifying a distinct lack of systems integrating Explainable AI (XAI) with Pharmacogenomic (PGx) guidelines. Addressing this gap requires both a scalable data engineering infrastructure capable of processing state-level claims data and a principled method for distinguishing causal drug exposures from correlated documentation artifacts. Chapter 2 presents two foundational contributions that together underpin all subsequent empirical chapters. First, it details the partition-first APCD data architecture — processing 15 Age Band × Cohort partitions via parallel DuckDB workers with strict temporal leakage prevention. Second, it introduces the **Consensus Filter** (SHAP ∩ FFA): a feature is designated causal only when independently confirmed by both distributional (SHAP) and structural (Formal Feature Attribution, FFA) analysis, directly addressing the failure modes of each method used alone. Together, these establish the replicable infrastructure and causal feature selection methodology applied throughout Chapters 3, 4, and 5.

*Chapter 2 is currently prepared for submission to CPT: Pharmacometrics & Systems Pharmacology and has been formatted to that journal's style.*

---

```{r ch02}
#| results: asis
include_chapter("../CH_2/ch02_psp.qmd")
```

\cleardoublepage

# Chapter 3: Causal Temporal Drivers of Opioid-Related Emergency Department Visits in Young and Mid-Life Adults: An Ensemble Machine Learning Study with Consensus-Based Feature Attribution {#chap-3}

**Foreword to Chapter 3**

The previous method chapter provided a comprehensive overview of the partition-first data architecture and the strict target leakage prevention protocols designed to isolate true predictive features. Utilizing this validated pipeline, Chapter 3 applies these methodological constraints to the first specific clinical challenge: high-risk Emergency Department (ED) utilization among patients with Opioid Use Disorder (OUD). This chapter explores how combining exploratory sequence mapping with an Explainable AI "Consensus Filter" can successfully translate complex claims data into verifiable clinical pathways.

*Chapter 3 is currently prepared for submission to Clinical and Translational Science and has been formatted to that journal's style.*

---

```{r ch03}
#| results: asis
include_chapter("../CH_3/ch03_cts.qmd")
```

\cleardoublepage

# Chapter 4: Formal Feature Attribution as a Causal Calculator for Drug-Drug Interaction Risk in Polypharmacy: A Systems Pharmacology Approach Using All-Payer Claims Data {#chap-4}

**Foreword to Chapter 4**

The previous results chapter examined the application of our Explainable AI pipeline to predict opioid-related ED visits, successfully extracting verifiable drivers from tree-based machine learning models. Chapter 4 follows on to adapt this methodology for a different clinical crisis: preventable ED visits driven by geriatric polypharmacy. By intentionally restricting the predictive feature space strictly to drug exposures and CPIC counts, this chapter demonstrates how the pipeline functions as a dedicated "causal calculator" via Formal Feature Attribution (FFA) to discover synergistic and antagonistic drug-drug interactions.

*Chapter 4 is currently prepared for submission to CPT: Pharmacometrics & Systems Pharmacology and has been formatted to that journal's style.*

---

```{r ch04}
#| results: asis
include_chapter("../CH_4/ch04_psp.qmd")
```

\cleardoublepage

# Chapter 5: A Privacy-First Serverless Pharmacogenomic Risk Dashboard: Translating Ensemble Models and Causal Rules to Clinical Decision Support {#chap-5}

**Foreword to Chapter 5**

The previous chapters successfully developed and applied a robust predictive pipeline, generating verifiable causal rules for both opioid and polypharmacy cohorts. However, identifying causal rules is only one part of the solution; these insights must be translated into actionable clinical practice. Chapter 5 follows on to present the final technological implementation of this research: translating the ensemble models and Pharmacogenomic (PGx) guidelines into a secure, serverless AWS clinical decision support dashboard. This chapter explores the system architecture required to safely match patient genotypes against predictive risk models in real-time.

*Chapter 5 is currently prepared for submission to the Journal of Personalized Medicine and has been formatted to that journal's style.*

---

```{r ch05}
#| results: asis
include_chapter("../CH_5/ch05_cpt.qmd")
```

\cleardoublepage

# Chapter 6: Synthesis and Future Directions {#chap-6}

**Foreword to Chapter 6**

The previous chapters addressed the specific technical and clinical challenges of predictive modeling, from raw data engineering (Chapter 2) and causal rule extraction (Chapters 3 and 4) to real-world dashboard deployment (Chapter 5). Chapter 6 serves as the final synthesis of this dissertation. It draws together the main findings of the preceding five manuscripts, explicitly arguing the coherence of the work as a whole and evaluating how these combined efforts answer the overarching research aims of advancing clinical predictive modeling via Explainable AI and precision therapeutics.

---

```{r ch06}
#| results: asis
include_chapter("../CH_6/ch06_conclusion.qmd")
```

\cleardoublepage

# References {.unnumbered}

::: {#refs}
:::
